Conceptualizing open health data platforms for low- and middle income countries

### 2024-05-27 index.tex
wip

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Mark van der Graaf; Nicole Spieker; Mark de Reuver},
  pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
platforms, LMICs, digital health},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

@@

Case studies on digital platforms in healthcare point to an emerging
pattern where the focus shifts from the digital platform with its
defining software and hardware components, to the data as the primary
object of interest in and of itself
\citep{ozalp2022digital, alaimo2022organizations}. This observation ties
into the proposed research agenda by de Reuver et al.~to consider data
platforms as a phenomenon distinct from digital platforms
\citep{dereuver2022openness}. Generally, data platforms inherit the
characteristics of digital platforms. For example, the economic
perspective on digital platforms stresses their multi-sidedness with
developers and consumers, while data platforms are used by data owners,
data consumers and third party solution providers. At the same time,
data platforms differ as their main offerings revolve around data. From
a market perspective, data platforms have more moderate network effects
and are more susceptible to fragmentation and heterogeneity.

@@

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods \citep{digitalpublicgoods}. If we are to use the OpenHIE
framework as an open data platform, we need to extend the standards,
technologies and architecture to include functionality for data sharing
and reuse. The lack of detailed specifications and consensus of this
addition to OpenHIE currently stands in the way of development projects
that aim to establish HDPs in LMICs. The purpose of this paper,
therefore, is to specificy how new standards and technologies can be
integrated into the OpenHIE architecture such that an open HDP can be
realized that supports the four types of data sharing as defined by de
Reuver at al.~(Table~\ref{tbl-types-data-sharing}).

@@

Data management and analytics platforms have undergone significant
changes since the first generation of data warehouses were introduced.
Recent studies have shown that the current practice has converged
towards the lakehouse as one of the most commonly used solution designs
\citep{armbrust2021lakehouse, hai2023data, harby2022data}. Lakehouses
typically have a zonal architecture that follow the
Extract-Load-Transform pattern (ELT) where data is ingested from the
source systems in bulk (E), delivered to storage with aligned schemas
(L) and transformed into a format ready for analysis (T)
\citep{hai2023data}. The discerning characteristic of the lakehouse
architecture is its foundation on low-cost and directly-accessible
storage that also provides traditional analytical DBMS management and
performance features such as ACID transactions, data versioning,
auditing, indexing, caching, and query optimization
\citep{armbrust2021lakehouse}. Lakehouses thus combine the key benefits
of data lakes and data warehouses: low-cost storage in an open format
accessible by a variety of systems from the former, and powerful
management and optimization features from the latter.

@@

While the lakehouse architecture separates the concerns of compute and
storage, the composable data stack takes the separation of concerns is
taken one step further. A composable data system
(Figure~\ref{fig-composable-data-stack}), not only separates the storage
(layer 3) and execution (layer 2), but also separates the user interface
(layer 1) from the execution engine by introducing standards including
Substrait for Intermediate Representation (standard A, IR) and Apache
Arrow for data connectivity and data memory layout (standards B and C,
respectively). The composable data stack can be implemented with current
open source technologies (Figure~\ref{fig-cds-examples}). As an example,
the Ibis user interface is currently sufficiently mature to offer a
standardized dataframe interface to 19 different execution engines.

@@

\includegraphics{images/composable-data-stack.png}

@@

\caption{\label{fig-composable-data-stack}The composable data stack.
Image source: \url{https://voltrondata.com/codex}.}

@@

\includegraphics{images/composable-data-stack-implementation-2.png}

@@

\caption{\label{fig-cds-examples}Examples of implementations of the
composable data stack. Image source:
\url{https://voltrondata.com/codex}.}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2558}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2558}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4884}}@{}}
\caption{Overview of current open source implementations of components
included in the OpenHIE specification that are FHIR-compatible. The
category Analytics Services is not a part of the original OpenHIE and is
discussed in the paper. Point-of-Service systems are excluded for
brevity. A systematic review of such digital public goods is beyond the
scope of this document.}\label{tbl-digital-public-goods}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Component
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Digital Public Good
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Component
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Digital Public Good
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Interoperability layer (IOL)} & IOL &
\href{http://openhim.org}{OpenHIM} \\
\textbf{Registry Services} & Client Registry (CR) &
\href{https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi}{SanteMPI} \\
& & \href{https://jembi.gitbook.io/jempi/}{JeMPI} \\
& & \href{https://www.openclientregistry.org/}{OpenCR} \\
& Facility Registry (FR) &
\href{https://github.com/intrahealth/gofr}{Global Open Facility Registry
(GOFR)} \\
& Health Worker Registry (HWR) &
\href{https://www.ihris.org/ihris-50}{iHRIS} \\
& Terminology Service (TS) &
\href{https://openconceptlab.org/terminology-service/}{OCL Terminology
Service} \\
& Product Catalog (PC) & not available \\
\textbf{Business Domain Services} & Shared Health Record (SHR) &
\href{https://hapifhir.io/}{HAPI FHIR} \\
& & \href{https://www.health-samurai.io/fhirbase}{Fhirbase} \\
& & \href{https://github.com/microsoft/fhir-server}{FHIR Server for
Azure} \\
& & \href{https://github.com/I-TECH-UW/shared-health-record}{I-TECH-UW
SHR} \\
& Health Management Information System (HMIS) &
\href{http://dhis2.org/}{DHIS2} \\
& Finance and Insurance Service (FIS) &
\href{http://openimis.org/}{OpenIMIS} \\
& Logistics Management Information System (LMIS) &
\href{http://openlmis.org/}{OpenLMIS} \\
\textbf{Generic} & Identity and Access Management (IAM) &
\href{https://www.keycloak.org/}{Keycloak} \\
& Gateway \& proxy & \href{https://github.com/google/fhir-gateway}{FHIR
Information Gateway} \\
& Admin dashboard for SHR &
\href{https://github.com/onaio/fhir-web}{OpenSRP FHIR Web} \\
& Configuration and deployment &
\href{https://jembi.gitbook.io/instant-v2/}{Instant OpenHIE} \\
\end{longtable}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CLI & Command-line Interface \\
CR & Client Registry \\
ELK & Elasticsearch, Logstach and Kibana stack \\
ELT & Extract, Load and Transform \\
FAIR & Findable, Accessible, Interoperable and Reusable \\
FHIR & Fast Healthcare Interoperability Resources \\
FL & Federated learning \\
HDP & Health data platform, explicitly differentiated from health
digital platform \\
HIE & Health Information Exchange \\
IR & Intermediate Representation \\
LMIC & Low- and middle income countries \\
MPC & Multiparty Computation \\
PET & Privacy-enhancing technologies \\
SHR & Shared Health Record \\
\end{longtable}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Conceptualizing open health data platforms for low- and middle income countries},
  pdfauthor={Daniel Kapitan; Femke Heddema; Julie Fleischer; Chris Ihure; Antragama Abbas; Steven Wanyee; Alessandro Pietrobon; Ryan Chrichton; John Grimes; Paula van Brakel; Mark van der Graaf; Nicole Spieker; Mark de Reuver},
  pdfkeywords={Analytics-on-FHIR, SQL-on-FHIR, HIE, data
platforms, LMICs, digital health},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\author[1]{Paula van Brakel%
%
}

@@

Case studies on digital platforms in healthcare point to an emerging
pattern where the focus shifts from the digital platform with its
defining software and hardware components, to the data as the primary
object of interest in and of itself
\citep{ozalp2022digital, alaimo2022organizations}. This development
roughly correlates with the commonly used distinction between primary
and secondary health data sharing \citep{cascini2024health}. While
people are generally willing to share their health data to receive the
best care (primary use), the attitude towards secondary use of health
data varies greatly depending on the type and context. The shift in
perspective from digital platforms supporting primary data sharing
toward data platforms supporting secondary data sharing, ties in with
the proposed research agenda by de Reuver et al.~to consider data
platforms as a phenomenon distinct from digital platforms
\citep{dereuver2022openness}. Generally, data platforms inherit the
characteristics of digital platforms. From an economic perspective, for
example, both types exhibit multi-sided markets. At the same time, data
platforms differ as their main offerings revolve around data. From an
ecosystem perspective, data platforms have more moderate network effects
and are more susceptible to fragmentation and heterogeneity.

@@

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods \citep{digitalpublicgoods}. Typically, these open source
components are intended to support deployment in small countries
(population up to 10 million) or large NGOs out of the box, and should
provide a stepping stone for customized deployments in medium-sized
countries (population around 40 million).\footnote{Although the OpenHIE
  specification does not include details on dimensioning, these are
  typically the requirements that are used within the community. See
  \href{https://wiki.ohie.org/display/CP/Interoperability+Layer+-+Use+Cases+and+Requirements}{OpenHIE
  Community Wiki}.} If we are to use the OpenHIE framework as an open
data platform, we need to extend the standards, technologies and
architecture to include functionality for data sharing and reuse. The
lack of detailed specifications and consensus of this addition to
OpenHIE currently stands in the way of development projects that aim to
establish HDPs in LMICs. The purpose of this paper, therefore, is to
specificy how new standards and technologies can be integrated into the
OpenHIE architecture such that an open HDP can be realized that supports
the four types of data sharing as defined by de Reuver at
al.~(Table~\ref{tbl-types-data-sharing}).

@@

Data management and analytics platforms have undergone significant
changes since the first generation of data warehouses were introduced.
Recent studies have shown that the current practice has converged
towards the lakehouse as one of the most commonly used solution designs
\citep{armbrust2021lakehouse, hai2023data, harby2022data}. Lakehouses
typically have a zonal architecture that follow the
Extract-Load-Transform pattern (ELT) where data is ingested from the
source systems in bulk (E), delivered to storage with aligned schemas
(L) and transformed into a format ready for analysis (T)
\citep{hai2023data}. The discerning characteristic of the lakehouse
architecture is its foundation on low-cost and directly-accessible
storage that also provides traditional database management and
performance features such as ACID transactions, data versioning,
auditing, indexing, caching, and query optimization
\citep{armbrust2021lakehouse}. Lakehouses thus combine the key benefits
of data lakes and data warehouses: low-cost storage in an open format
accessible by a variety of systems from the former, and powerful
management and optimization features from the latter.

@@

While the lakehouse architecture separates the concerns of compute and
storage, the composable data stack takes the separation of concerns is
taken one step further. A composable data system
(Figure~\ref{fig-composable-data-stack}), not only separates the storage
(layer 3) and execution (layer 2), but also separates the user interface
(layer 1) from the execution engine by introducing standards including
Substrait for Intermediate Representation (standard A, IR) and Apache
Arrow for data connectivity and data memory layout (standards B and C,
respectively). The composable data stack can be implemented with current
open source technologies (\textbf{?@fig-cds-examples}). As an example,
the Ibis user interface is currently sufficiently mature to offer a
standardized dataframe interface to 19 different execution engines.

@@

\begin{minipage}{\linewidth}

\begin{figure}[H]

{\centering \includegraphics{images/composable-data-stack.png}

@@

\subcaption{a)}

@@

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\begin{figure}[H]

{\centering \includegraphics{images/composable-data-stack-implementation-2.png}

@@

\subcaption{b)}

\end{minipage}%

\caption{\label{fig-composable-data-stack}Schematic overview of the
composable data stack showing a) the overall architecture and b)
examples of implementations. Images taken from
\url{https://voltrondata.com/codex}.}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2558}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2558}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4884}}@{}}
\caption{Overview of current open source implementations of components
that fit in the OpenHIE specification. The category Analytics Services
is not a part of the original OpenHIE and is discussed in this paper.
Point-of-Service systems are excluded for brevity.List compiled using
\href{https://wiki.ohie.org/display/DR/OpenHIE+Reference+Technologies}{OpenHIE
Reference Technologies},
\href{https://digitalsquare.org/digital-health-global-goods}{Global
Goods for Digital Health},
\href{https://digitalpublicgoods.net/}{Digital Public Goods Alliance}
and search of open source code repositories. A systematic review of such
digital public goods is beyond the scope of this
document.}\label{tbl-digital-public-goods}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Component
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Digital Public Good
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Component
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Digital Public Good
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Interoperability layer (IOL)} & IOL &
\href{http://openhim.org}{OpenHIM} \\
& & \href{https://www.mhero.org/technology}{mHero} \\
& & \href{https://docs.openfn.org/}{OpenFN} \\
\textbf{Registry Services} & Client Registry (CR) &
\href{https://help.santesuite.org/product-overview/santesuite-products/master-patient-index-santempi}{SanteMPI} \\
& & \href{https://jembi.gitbook.io/jempi/}{JeMPI} \\
& & \href{https://www.openclientregistry.org/}{OpenCR} \\
& & \href{https://www.opencrvs.org/}{OpenCRVS} \\
& Facility Registry (FR) &
\href{https://github.com/intrahealth/gofr}{Global Open Facility Registry
(GOFR)} \\
& & \href{https://geoprismregistry.com/}{GeoPrism Registry (GPR)} \\
& & \href{http://instedd.org/technologies/resource-map/}{Resource
Map} \\
& & \href{https://healthsites.io/}{Healthsite.io} \\
& & \href{https://geoprismregistry.com/}{GeoPrism Registry} \\
& & \href{https://github.com/dhis2}{DHIS2} \\
& Health Worker Registry (HWR) &
\href{https://www.ihris.org/ihris-50}{iHRIS} \\
& Terminology Service (TS) &
\href{https://openconceptlab.org/terminology-service/}{OCL Terminology
Service} \\
& Product Catalog (PC) & \href{https://productcatalog.io/}{PCMT} \\
\textbf{Business Domain Services} & Shared Health Record (SHR) &
\href{https://hapifhir.io/}{HAPI FHIR} \\
& & \href{https://www.health-samurai.io/fhirbase}{Fhirbase} \\
& & \href{https://github.com/microsoft/fhir-server}{FHIR Server for
Azure} \\
& & \href{https://github.com/I-TECH-UW/shared-health-record}{I-TECH-UW
SHR} \\
& Health Management Information System (HMIS)\footnote{Note that there
  is often a confusion on the acronym HMIS. Strictly speaking, the
  OpenHIE specification uses HMIS to refer to a Health Management
  Information System that is part of the Business Domain Services.
  Sometimes HMIS is used to refer to a Hospital Management Systems in
  the Point-of-Service domain, synonymous with an Electronic Medical
  Record (EMR) system.} & \href{http://dhis2.org/}{DHIS2} \\
& Finance and Insurance Service (FIS) &
\href{http://openimis.org/}{OpenIMIS} \\
& Logistics Management Information System (LMIS) &
\href{http://openlmis.org/}{OpenLMIS} \\
& & \href{https://openboxes.com/features/}{OpenBoxes} \\
\textbf{Generic} & Identity and Access Management (IAM) &
\href{https://www.keycloak.org/}{Keycloak} \\
& Gateway \& proxy & \href{https://github.com/google/fhir-gateway}{FHIR
Information Gateway} \\
& Admin dashboard for SHR &
\href{https://github.com/onaio/fhir-web}{OpenSRP FHIR Web} \\
& Configuration and deployment &
\href{https://jembi.gitbook.io/instant-v2/}{Instant OpenHIE} \\
\end{longtable}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5000}}@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ACID & Atomicity, Consistency, Isolation, and Durability \\
CLI & Command-line Interface \\
CR & Client Registry \\
ELK & Elasticsearch, Logstach and Kibana stack \\
ELT & Extract, Load and Transform \\
FAIR & Findable, Accessible, Interoperable and Reusable \\
FHIR & Fast Healthcare Interoperability Resources \\
FL & Federated learning \\
HDP & Health data platform, explicitly differentiated from health
digital platform \\
HIE & Health Information Exchange \\
IR & Intermediate Representation \\
LMIC & Low- and middle income countries \\
MPC & Multiparty Computation \\
PET & Privacy-enhancing technologies \\
SHR & Shared Health Record \\
\end{longtable}
